Static contract checking for Haskell

### 2011-12-20 paper/paper.tex
Paper: removed most mention of old translation.
The fancy figures still have the old trans.  The plain text figures
use the new translation.

The syntax of $\hprime$ is defined in figure~\ref{hprime-stx}. A
module is a list of toplevel definitions, claims that functions
statisfy contracts and data definitions.  The $\hprime$ language we
present is minimalistic, but the implementation supports a little
more.

@@

\begin{figure}[h]
  \centering
  \[
  \begin{array}{rclr}
    mod &:=& def_1,\dots,def_n &\\
    def &\in& \mbox{Definition} & \\
    def &:=& \texttt{data } T = K_1 a_1 \mid \cdots \mid K_n a_n& \mbox{Data definition}\\
    &\mid& f \in c & \mbox{Contract claim}\\
    &\mid& f~\vec{x}~=~e & \\
    &\raisebox{ 0.18 cm }{$\mid$}& \multicolumn{2}{c}{
      \begin{array}{rcl}
        \hspace{-0.2cm} f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
         && K_1~\vec{x}_1 \to e_1 \mid \cdots \mid K_n~\vec{x}_n \to e_n \\
       \end{array}
       } \\
  \end{array} \]

@@

\caption{Syntax of the language $\hprime$}
\label{hprime-stx}
\end{figure}

\subsection{Contracts}
Contract syntax is described in figure \ref{cont-stx}. The predicates
we use in our contracts can be any boolean $\hprime$ expression. We
only consider pairs of contract for simplicity, although there is no
issue with generalisation to arbitrary tuples.\note[nc]{
%
  XXX: Tuple contracts are not actually supported in the
  implementation. Should they be discussed here?
%
So, ``tuples'' refers to the the term constructor, not the type
constructor, and the generalization is e.g.
\[
e \in Cons~c_1~c_2 \iff e \text{ diverges or } 
      (e \to^\star Cons~e_1~e_2 \mbox{ and } e_1 \in c_1, e_2 \in c_2)
\]
?
}

@@

The notation $\bad \notin_s e$ means that $\bad$ does not appear
anywhere in $e$ or $e$'s (transitive) dependencies.  For
example, $Just ~3$ is syntactically safe whereas $Just ~\bad$ is not.

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.\note[nc]{I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e. forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.} The expression language in our logic is \hprime.

@@

\subsection{$\ctrans{}$ -- Contracts}\label{sec:contracts}
\note[nc]{%
Figure~\ref{fig:newCTrans} contains the new min contract translation, and
Figure~\ref{fig:naiveCTrans} contains the naive contract translation
with no $\min$'s.  The min translation reduces to the naive translation when
all $\min$'s are assumed true, i.e. when an axiom $\forall x.\min{x}$ is added.

The min translation has two forms, $\ctrans{}^-$ for
assumptions/axioms and $\ctrans{}^+$
for conclusions/theorems.  The choice of minus and 
plus comes from the negative and positive
positions in an implication: $- \to +$.  The goal of the $\min$'s is to limit the
applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ttrans{}$), 
in order to prune 
the search space.  So, $\min$'s
appear as obligations (the premises in implications) 
in contract and definition axioms. The mnemonic is ``(a)m in(terested)'',
and the translation is designed so that $\min{e}$ is introduced whenever
$e$ is evaluated or inspected.  The axiom translations contain $\min{}$
obligations that prevent using the axioms on ``uninteresting'' terms.

@@

As another example, consider the $\dtrans{}$ for functions defined by pattern matching,
in Figure~\ref{dtrans}.  Definition translations are axioms, 
so the the initial $\min{\full{f}(\vec{x})}$
is an obligation: we aren't allowed to 
``unfold the definition'' of the function unless we prove
that we're interested in a particular application.  
Once that obligation is satisfied, interest ($\min{e}$) in the
scrutinee of the case match may be concluded, because the function
must inspect the scrutinee to compute the result.

Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
Figure~\ref{fig:newCTrans}.  To prove that $e \in \{x|p\}$, we get to
assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
we need to prove we're interested in $e$.  The $\min{e}$ constraint
here is not optional, because in addition to concluding 
$e \neq \unr \to p[e/x] \in \{\true,\unr\}$, we also conclude $\min{p[e/x]}$.
The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
$p[e/x] \in \{\true,\unr\}$, and because it can be used to unfold definitions,
we guard it with $\min{e}$.
}

@@

\begin{figure*}
\begin{verbatim}
[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
[[e in {x|p}]]  := e /= unr -> p[e/x] in {true, unr}

[[e in cf]]+ :=  min(e) ->  cf(e)
[[e in cf]]- := [min(e) ->] cf(e)

-- 'min's optional if a
-- 'forall f,x. min(f x) -> min(f)' axiom
-- is added to prelude.
[[e in x:c1 -> c2]]+
  := forall x. [min(e x) ->] [[x in c1]]- -> [[e x in c2]]+   
[[e in x:c1 -> c2]]-
  := forall x. [min(e x) ->] [[x in c1]]+ -> [[e x in c2]]-
\end{verbatim}
\caption{$\ctrans{}$ -- New contract translation. Design choices, in the form
of optional parts, are enclosed in single square brackets.}
\label{fig:newCTrans}
\end{figure*}

We give in figure~\ref{ctrans} the translation of contract satisfaction.
\remove[nc]{Note that we define the translation of $f \in c$ and of $f \not \in
c$. We have to do that because one is not the negation of the other,
even though $\lnot (f \not \in c)$ implies $f \in c$.}\note[nc]{

The contract translation, $\ctrans{}$, introduces a new predicate,
$\min{}$, and contract non-satisfaction, $f \notin c$.  The formula
$\min{e}$, short for ``(a)m in(terested) in $e$'' (?), is used to
guide the proof search, by restricting proofs to follow the evaluation
order when instantiating hypotheses (Is this meaningful? Is it even
correct?).  To understand the translation intuitively, simply ignore
all $\min{}$'s (I.e., assume $\forall x. \min{x}$).  The
non-satisfaction form, $f \notin c$, is introduced to make the
$\min{}$'s appear positively in translations in negative/contravariant
positions; Ignoring $\min{}$'s, we have $\ctrans{f \notin c} \equiv
\lnot \ctrans{f \in c}.$ (Well, after I change the definitions of
$\ctrans{}$ for $e \in \{x | p\}$ and $e \notin \{x|p\}$ \ldots)

Dimitrios gave an example where defining $\ctrans{f \notin c}$ as
$\lnot \ctrans{f \in c}$ is problematic: prove $\ctrans{g ~Z \in
  \cfc}$, assuming $\min{g ~Z}$, for an abstract function $g \in \cfc
\to \cfc$.  If we define 
\protect \[ 
\ctrans{g \in x:c1 \to c2} :=
\forall x. \min{g x} \to \ctrans{x \in c1} \to \ctrans{g x \in c2},
\]
then we get stuck, because we must prove $\min{Z}$ to satisfy
$\ctrans{x \in c1}$ (i.e. $\min{Z} \land \cf{Z}$; we have $\cf{Z}$ by
definition).  (UPDATE: Actually, this is probably just more evidence
that $\ctrans{e \in \cfc}$ is defined incorrectly: we were already
worried about that $\min{}$, because it stops us from proving
$\cf{Z}$, even when no arrows are involved).  On the other hand, if we
define the arrow case as in Figure~\ref{ctrans}, namely
\protect \[
\ctrans{g \in x:c1 \to c2} 
:= \forall x. \min{g x} \to \lnot \ctrans{x \notin c1} \to \ctrans{g x \in c2},
\]
then we succeed, because $\min{Z}$ is irrelevant to us in satisfying
$\ctrans{x \in c1}$ (i.e. $\lnot \min{Z} \lor \cf{Z}$. UPDATE: when you
notice that that implies $\min{Z} \to \cf{Z}$, (and the reverse follows from EM), the next 
paragraph might become a little dubious.  I still think the second form 
is more intuitive.).

The irrelevance of $\lnot \min{Z}$ above leads to the question, can we
do away with the $\min{}$'s altogether in such positions?  It turns
out the answer is ``No'', when we consider proving $g ~(g ~Z) \in
\cfc$, but I have some ideas for an alternative translation where the
answer is ``Yes''.  But first $g ~(g ~Z)$: to use $g$'s contract
assumption we need to prove $\lnot \min{g ~Z} \lor \cf{g ~Z}$, and to
prove $\cf{g ~Z}$ we in turn need to use $g$'s contract again.  To use
$g$'s contract again we need $\min{g ~Z}$, and so $\min{}$ is relevant
again.  The solution is to invoke excluded middle on $\min{g ~Z}$:
either $\min{g ~Z}$, and we're done, or $\lnot \min{g ~Z}$, in which
case we satisfy $g$'s preconditions \emph{without} showing $g$'s
argument to be crash free.  This seems a little surprising to me,
making a translation with no $\min{}$ here even more appealing.

Here's a start at such a translation.  I
think it's more natural to think about variance, so write
$\ctrans{}^+$ for conclusions and $\ctrans{}^-$ for assumptions.  Then
a system that works for these $\cfc$ examples is
\protect \begin{eqnarray*}
\ctrans{x \in \cfc}^+ &:=& \min{x} \to \cf{x} \\
\ctrans{x \in \cfc}^- &:=& \cf{x}\\
\ctrans{f \in x:c1 \to c2}^+ &:=& \forall x.               
  \ctrans{x \in c1}^- \to \ctrans{f x \in c2}^+\\
\ctrans{f \in x:c1 \to c2}^- &:=& \forall x. \min{f x} \to 
  \ctrans{x \in c1}^+ \to \ctrans{f x \in c2}^-
\end{eqnarray*}
Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
it's natural to think of contract assumptions in the theory as
negative, and contract goals as positive.  An example is helpful:
\protect \begin{eqnarray*}
\ctrans{g \in \cfc \to \cfc}^+ &=& \forall x. \cf{x} \to (\min{g ~x} \to \cf{g ~x}),\\
&&\text{whereas}\\
\ctrans{g \in \cfc \to \cfc}^- &=& \forall x. \min{g ~x} \to (\min{x} \to \cf{x}) \to \cf{g ~x}.
\end{eqnarray*}
Now, given $\ctrans{g \in \cfc \to \cfc}^-$, we may 
want to prove $\ctrans{g ~(g ~Z) \in \cfc}^+$.
So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
This time the proof proceeds more naturally (no EM).  
From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
$\min{g ~Z} \to \cf{g ~Z}$ (UPDATE: this is actually the same as 
before, up to using EM to rewrite the implication).
But that follows by instantiating $g$'s contract again, and 
then using $\cf{Z}$ to get $\min{Z} \to \cf{Z}$.

Notice the assymetry between the positive and negative arrow
translations: the positive case has no $\min{f ~x}$ assumption.  I
left this $\min{}$ out because it's already in the 
conclusion of the $\cf$ example, but I certainly haven't
worked enough examples to justify it.  Moreover, this fragment
(no $e \in \{x|p\}$ yet) already fails to have a standard
property of bidirectional systems: that you can check/verify what you assume/use.
In other words, we can't prove that, forall $e$ and $c$,
\protect \[
  \ctrans{e \in c}^- \text{ implies } \ctrans{e \in c}^+.
\]
The trouble seems to come down to needing to prove $\min{e ~x}$ implies 
$\min{e}$, which is not true a priori.  But, it's easy to add the $\min{g ~x}$ 
assumption to the positive arrow translation to recover the
 ``negative implies positive'' property. (UPDATE: the symmetric $\min{}$
may also be necessary when reasoning about functions that return functions: if
$f ~x = g$ then $f ~x ~y = g ~y$, but we'd need a $\min{f ~x}$ to unfold $f$'s def).

Going back to the current translation, we consider some of its
metatheoretic properties.
\protect \begin{enumerate}
\item $\lnot \ctrans{e \notin c}$ should behave like a double-negation
  translation, in the sense that $\ctrans{e \in c}$ should imply
  $\lnot \ctrans{e \notin c}$.  This becomes true if we modify the
  predicate case so that the implication holds (e.g. by more accurately
  modeling the operation semantics of contracts), and add a symmetric 
  $\min{f ~x}$ assumption to $\ctrans{f \notin x:c_1 \to c_2}$.  Note that
  reverse implication does not hold, e.g. $\lnot \ctrans{x \in \cfc}$ does
  not imply $\ctrans{x \in \cfc}$.
\item The ``fancy'' $\min{}$-ified translation should 
  degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
  This is trivially true if we define the ``naive'' translatino as the ``min'' tranlation
  plus that axiom.  In symbols, we have
  \protect \[
  \Gamma^\text{min} \vdash \phi \implies \Gamma^\text{naive} \vdash \phi.
  \]  I.e., ``\min{} is sound for `naive'''.
\item We'd like to have a result  like: given some validity or consistency 
  assumptions on $\Gamma$ and $e \in c$,
  \protect \[
    \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
  \]  I.e., ``\min{} is complete for `naive', under some assumptions about 
  what we're proving''.  However, Dimitrios points out that we shouldn't expect this
  completeness theorem without some work, since we're purposefully 
  restricting the proof search to follow
  evaulation strategy, and it's far from obvious that all proofs can proceed this way.

\end{enumerate}

Concerns about the current translation
\protect \begin{enumerate}
\item The changes needed to make the first property above hold (a $\min{}$ in the negative
  arrow case and symmetric predicate contract cases).
  I had trouble proving
  \protect \[
  (\lambda x. \text{case $x$ of $\true \to \true$; $\false \to \bad$})
  \in \{x|x==\true\} \to \cfc
  \]
  , after adding $p = \unr$ to $\ctrans{x \in \{x|p\}}$, but I need to think about it more.

\item There may need to be some $\min{}$ assumptions in the prelude.  E.g. right now 
  it's impossble to prove $\ctrans{\true \in \cfc}$, because you can't prove
  $\min{\true}$ a priori.  On the other hand, we don't need any extra
  $\min{}$ assumption in the arrow case, since it's already assumed, at least in
  very simple examples (checked on the simple
  example that $id \in \cfc \to \cfc$). (UPDATE: removing the $\min{x} \land$ in $cfc$, 
  or changing it to $\min{x} \to$, is probably the solution).

\item A related question, is what do we feed to Equinox?  Equinox is a 
  refutation prover, so to prove $\Gamma \vdash \phi$ we ask if 
  $\Gamma,\lnot \phi \vdash \bot$.  So, for us, when $\phi$ is $\ctrans{e \in c}$, do
  give Equinox $\Gamma,\lnot \ctrans{e \in c}$ or $\Gamma, \ctrans{e \notin c}$.  I think
  we want the former, since ignoring Equinox our goal is $\Gamma \vdash \ctrans{e \in c}$.
\end{enumerate}
}

\note[nc]{Here's a more fleshed out alternate translation, and a
  comparison with the current translation.}
\small
\begin{verbatim}
new translation
===============

-- you can assume both min()s when proving e in {x|p} 
[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
-- you need to prove min(e) but then get min(p[e/x])
[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
-- could also write this as 'e /= unr -> p[e/x] in {true,unr}'
[[e in {x|p}]]  := e = unr \/ p[e/x] in {true, unr}

[[e in cf]]+ := min(e) -> cf(e)
-- i think we can add a 'min(e) ->' here without losing anything.
[[e in cf]]- :=           cf(e)

-- should add a 'min(e x) ->' here as discussed above.
[[e in x:c1 -> c2]]+
  := forall x.             [[x in c1]]- -> [[e x in c2]]+   
[[e in x:c1 -> c2]]-
  := forall x. min(e x) -> [[x in c1]]+ -> [[e x in c2]]-

old translation (written in implication style)
==============================================

[[e in {x|p}]]       
  := min(p[e/x]) /\ (e = unr \/ p[e/x] = true)
not[[e notin {x|p}]] 
  := min(p[e/x]) -> (e /= bad /\ p[e/x] /= false)

[[e in cf]]       := min(e) /\ cf(e)
not[[e notin cf]] := min(e) -> cf(e)

[[e in x:c1 -> c2]] 
  := forall x. min(e x) -> not[[x notin c1]] -> [[e x in c2]]
not[[e notin x:c1 -> c2]]
  := forall x.             [[x in c1]] -> not[[e x notin c2]]

updated cf and {x|p} rules for old translation
==============================================

[[e in {x|p}]]       
  := min(p[e/x]) /\ (e = unr \/ p[e/x] in {true,unr})
not[[e notin {x|p}]] 
  := min(p[e/x]) -> (e = unr \/ p[e/x] in {true,unr})

-- what we actually discussed was 
-- adding a top level min(e) assumption, but
-- removing the min(e) goal has a similar effect.
[[e in cf]]       :=           cf(e)
not[[e notin cf]] := min(e) -> cf(e)
\end{verbatim}
\note[nc]{And what do we see?
If we identify $\in^+$ with $\lnot\notin$ and $\in^-$ with $\in$, 
then the ``new'' translation looks almost the same,
except it has more $\min{}$'s in the predicate rules.  So, have we learned
anything? Probably:
\protect \begin{enumerate}
  \item We're more confident about our suggested changes to the old translation,
        including adding a symmetric $\min{}$ assumption to the $\notin$ arrow case. 
  \item The new translation is more intuitive / we might understand how Koen came up
        with the old translation.
  \item We can answer the question about what we give as input to Equinox: for the
        new translation, it's obvious that the theory $T$ should consist of $\in^-$ axioms.
        That is consistent with giving Equinox $\in$ axioms.  
        What to give as the goal is maybe
        less obvious, since Equinox is a refutation prover, but consider: ignoring the
        strategy (refutation), we want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
        But
        that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.  
        And going back
        to the old translation, the analogy is 
        $\Gamma^{\in}, \ctrans{e \notin c} \vdash \bot$.
\end{enumerate}

The predicate case for the new translation came from an example:
}
\begin{verbatim}
try:

[[e in {x|p} ]]+ = min(e) -> min(p[e/x]) -> [[e in {x|p}]]-
[[e in {x|p} ]]- = e = unr \/ p[e/x] in {true, unr}

example:

e := (\x.case x of True -> True) -- so e false = bad
c := {x|x==True} -> CF
[[e in c]]+ = forall x. min(e x) 
              -> [[x in {x|x==True}]]- -> [[e x in CF]]+
            = forall x. min(e x) 
              -> (x = unr \/ p in {true,unr})
              -> (min(e x) -> CF(e x))

min(e x) -> min(x) /\ (
         \/ (x = bad /\ (e x) = bad) 
         \/ (x = True /\ (e x) = true) 
         \/ (x = False /\ (e x) = bad)
 -- better to add the redundant x /= bad, true, false here?
         \/ (e x) = unr
         )

if x = unr then (e x) = unr and so CF(unr) and done

if x /= unr /\ p = true then we're stuck, since we can't 
unroll the definition of ==true in p

so, change the defs:

[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
[[e in {x|p}]]  := e = unr \/ p[e/x] in {true, unr}

starting again, we have min(x) from D[[]] so we get

min(p) /\ (x = unr \/ p in {true,unr})

now

 - x = unr gives (e x) = unr as before

 - for x/=unr and p in {true,unr} we unroll def (==true) 
   using min(p), for p = ((==True) x):

   min((==True) x) -> min(x) /\ ( -- e' is (==True) here
         \/ (x = bad /\ (e' x) = bad) 
         \/ (x = True /\ (e' x) = true) 
         \/ (x = False /\ (e' x) = bad)
         \/ (e' x) = unr
         )

for p = unr, we get x notin {bad,true,false}, 
and hence (e x) = unr, and we're done.

for p = true, we get x = true, 
and hence (e x) = true, and we're done.
\end{verbatim}
\note[nc]{That example also has an alternate version of the
 definition translation for functions defined by case
matching.  It's a little more concise than the old translation.
}
\begin{verbatim}
In any case, another question: 
does equinox distinguish between D[[]] like

>>>>

The syntax of $\hprime$ is defined in figure~\ref{fig:hprime-syntax}. A
module is a list of toplevel definitions, claims that functions
statisfy contracts and data definitions.  The $\hprime$ language we
present is minimalistic, but the implementation supports a little
more.

@@

\begin{figure}[h]
  \centering
  \[
  \begin{array}{rclr}
    \hprime module &:=& def_1,\dots,def_n &\\
    def &\in& \mbox{Definition} & \\
    def &:=& \texttt{data } T = K_1 a_1 \mid \cdots \mid K_n a_n& \mbox{Data definition}\\
    &\mid& f \in c & \mbox{Contract claim}\\
    &\mid& f~\vec{x}~=~e & \\
    &\raisebox{ 0.18 cm }{$\mid$}& \multicolumn{2}{c}{
      \begin{array}{rcl}
        \hspace{-0.2cm} f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
         && K_1~\vec{x}_1 \to e_1 \mid \cdots \mid K_n~\vec{x}_n \to e_n \\
       \end{array}
       } \\
  \end{array} \]

@@

\caption{Syntax of the language $\hprime$}
\label{fig:hprime-syntax}
\end{figure}

\subsection{Contracts}
Contract syntax is described in figure \ref{cont-stx}. The predicates
$p$ in contracts range over boolean $\hprime$ expressions. We
only consider pairs of contract for simplicity, although there is no
issue with generalisation to arbitrary tuples.\note[nc]{
%
  XXX: Tuple contracts are not actually supported in the
  implementation. Should they be discussed here?
%
So, ``tuples'' refers to the the term constructor, not the type
constructor, and the generalization is e.g.
\[
e \in Cons~c_1~c_2 \iff e \text{ diverges or } 
      (e \to^\star Cons~e_1~e_2 \mbox{ and } e_1 \in c_1, e_2 \in c_2)
\]
?
}

@@

The notation $\bad \notin_s e$ means that $\bad$ does not appear
anywhere in $e$ or $e$'s (transitive) dependencies.  For
example, $Just ~3$ is syntactically safe whereas $Just ~jb$ is 
not, for $jb$ defined in a separate declaration as $jb = Just~\bad$.

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.\note[nc]{I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e. forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.} 
The expression language in our logic is the expression language of 
\hprime (Figure~\ref{fig:hprime-syntax}).  Note that curried application in \hprime
corresponds to a binary function symbol in FOL, which we denote by juxtaposition.

@@

\subsection{$\ctrans{}$ -- Contracts}\label{sec:contracts}
\note[nc]{%
Figure~\ref{fig:newCTrans} contains the $\min$-translation of contracts, and
Figure~\ref{fig:naiveCTrans} contains the naive contract translation,
with no $\min$'s.  The min translation reduces to the naive translation when
all $\min$'s are assumed true, i.e. when an axiom $\forall x.\min{x}$ is added.

The min translation has two forms, $\ctrans{}^-$ for
assumptions/axioms and $\ctrans{}^+$
for conclusions/theorems.  The choice of minus and 
plus comes from the negative and positive
positions in an implication: $- \to +$.  The goal of the $\min$'s twofold: 
(1) to allow finite counter models when the theory 
does not prove the contracts; (2) to limit the
applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ttrans{}$), 
in order to prune the proof search space.  The $\min$'s
appear as obligations (the premises in implications) 
in contract and definition axioms. The mnemonic is ``(a)m in(terested)'',
and the translation is designed so that $\min{e}$ is introduced whenever
$e$ is evaluated or inspected.  The axiom translations contain $\min{}$
obligations that prevent using the axioms on ``uninteresting'' terms.

@@

As another example, consider the $\dtrans{}$ for functions defined by pattern matching,
in Figure~\ref{dtrans}.  Definition translations are axioms, 
so the the initial $\min{\full{f}(\vec{x})}$
is an obligation: we aren't allowed to 
``unfold the definition'' of the function unless we prove
that we're interested in a particular application.  
Once that $\min$ obligation is satisfied, interest ($\min{e}$) in the
scrutinee of the case match may be concluded, because the function
must inspect the scrutinee to compute the result.

Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
Figure~\ref{fig:newCTrans}.  When proving that $e \in \{x|p\}$, we get to
assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
we need to prove we're interested in $e$.  The $\min{e}$ constraint
here is not optional, because in addition to concluding 
$e \neq \unr \to p[e/x] \in \{\true,\unr\}$, we also conclude $\min{p[e/x]}$.
The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
$p[e/x] \in \{\true,\unr\}$, and because it can be used to unfold definitions,
we guard it with $\min{e}$.
}

@@

\begin{figure*}
\begin{verbatim}
-- Need 'min(e)' and 'min(p[e/x])' here to unfold defs.
[| e ::: {x|p} |]+ = min(e) /\ min(p[e/x]) ->
                     (e /= unr -> p[e/x] not-in {false, bad})
-- Need 'min(e)' here to guard 'min(p[e/x])'.
-- Need 'min(p[e/x])' here to unfold defs.
[| e ::: {x|p} |]- = min(e) -> min(p[e/x]) /\
                     (e /= unr -> p[e/x]     in {true, unr})

 -- Need 'min(e)' here to unfold defs.
[[e in cf]]+ := min(e) ->  cf(e)
[[e in cf]]- := cf(e)

 -- No auxiliary mins.
[[e in x:c1 -> c2]]v
  := forall x. [[x in c1]](dual v) -> [[e x in c2]]v
\end{verbatim}
\caption{$\ctrans{}$ -- New contract translation. Design choices, in the form
of optional parts, are enclosed in single square brackets.}
\label{fig:newCTrans}
\end{figure*}

Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
it's natural to think of contract assumptions in the theory as
negative, and contract goals as positive.  An example is helpful:
\protect \begin{eqnarray*}
\ctrans{g \in \cfc \to \cfc}^+ &=& \forall x. \cf{x} \to (\min{g ~x} \to \cf{g ~x}),\\
&&\text{whereas}\\
\ctrans{g \in \cfc \to \cfc}^- &=& \forall x. \min{g ~x} \to (\min{x} \to \cf{x}) \to \cf{g ~x}.
\end{eqnarray*}
Now, given $\ctrans{g \in \cfc \to \cfc}^-$, we may 
want to prove $\ctrans{g ~(g ~Z) \in \cfc}^+$.
So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
$\min{g ~Z} \to \cf{g ~Z}$
But that follows by instantiating $g$'s contract again, and 
then using $\cf{Z}$ to get $\min{Z} \to \cf{Z}$.

Going back to the current translation, we consider some of its
metatheoretic properties.
\protect \begin{enumerate}
\item A standard
  property of bidirectional systems: that you can check/verify what you assume/use.
  In other words, forall $e$ and $c$,
  \protect \[
  \ctrans{e \in c}^- \text{ implies } \ctrans{e \in c}^+.
  \]
\item The ``fancy'' $\min{}$-ified translation should 
  degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
  This is trivially true if we define the ``naive'' translatino as the ``min'' translation
  plus that axiom.  In symbols, we have
  \protect \[
  \Gamma^\text{min} \vdash \phi \implies \Gamma^\text{naive} \vdash \phi.
  \]  I.e., ``\min{} is sound for `naive'''.
\item We'd like to have a result  like: given some validity or consistency 
  assumptions on $\Gamma$ and $e \in c$,
  \protect \[
    \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
  \]  I.e., ``\min{} is complete for `naive', under some assumptions about 
  what we're proving''.  However, Dimitrios points out that we shouldn't expect this
  completeness theorem without some work, since we're purposefully 
  restricting the proof search to follow
  evaulation strategy, and it's far from obvious that all proofs can proceed this way.
\end{enumerate}

  the theory $T$ should consist of $\in^-$ axioms.
  We want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
  But
  that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.

\begin{verbatim}
In any case, another question: 
does equinox distinguish between D[[]] like
